Examining equity in accessibility to bike share - a balanced floating catchment area approach

### 2021-06-21 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Edits after Elise revised the paper

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining equity in accessibility to bike share - a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{rm}\NormalTok{(}\DataTypeTok{list =} \KeywordTok{ls}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

@@

The potential of public bicycle share programs (PBSPs) to increase
bicycling levels is but one of many reasons for implementing such
programs in urban areas (see Hosford et al., 2018, 2019). As a healthy,
inexpensive, and convenient form of public transportation, shared
bicycles can encourage individuals to take up bicycling for short local
trips or first and last mile trips to other public transportation
instead of using personal vehicles. These programs can also play a role
in advancing transportation equity if they make bicycling more
accessible to disadvantaged populations. Although PBSPs are available to
the general public in over 800 cities worldwide (Fishman, 2016), and
ought to be accessible to any individual who wishes to use them,
research on PBSPs indicates that inequities persist with respect to who
can use and access them.

@@

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users
(Fishman, 2016). In many cities, males use bike share more than females
(Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
Reilly, S. M. Wang, et al., 2020; Winters et al., 2019) as do younger
age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011).
However, one study found that bike share users in Washington, DC were
more likely to be female (Buck et al., 2013), which suggests that the
gender gap among bicyclists who use PBSPs is less disparate than the gap
for personal bicycle use (Fishman, 2016). There is some evidence that
bike share users are less likely to own a car (Buck et al., 2013;
Reilly, Noyes, et al., 2020). However, the relationship between income
or education and bike share use is less clear-cut. Stations in
disadvantaged communities in Chicago have been found to generate most of
the average annual trips (Qian and Jaller, 2020) and individuals from
minority or lower socioeconomic status neighborhoods in
Minneapolis-St.~Paul also used the city's PBSP more (J. Wang and
Lindsey, 2019a). Similar findings were reported in London (Ogilvie and
Goodman, 2012). Being university educated was a significant correlate of
bike share use in Montreal, Canada (Fuller et al., 2011). Not
coincidentally, financial savings have been found to motivate those on a
low income to use bike share (Fishman, 2016).

@@

The introduction of PBSPs has been accompanied by a flurry of research
focusing on benefits from them. Many studies have examined differences
in demographics or socioeconomic status between those who use or have
access to PBSPs and those who don't, while other research has explored
spatial inequities in where stations are located (e.g., Hosford and
Winters, 2018; Hull Grasso et al., 2020; Mooney et al., 2019; Qian and
Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian and Jaller
(2020) estimated ridership in the city's PBSP and found that a minority
of bike share stations were located in disadvantaged communities, while
annual members from such areas had a lower share of trips compared to
other areas in the city. Similar results were found in Philadelphia.
Despite efforts to increase equity within the city's PBSP, census block
groups with lower median income generated fewer trips (Caspi and Noland,
2019). Trips from stations in such areas were utilitarian (e.g.,
commuting to work), which points to the importance of ensuring equitable
access (Caspi and Noland, 2019). In the case of Seattle, all
neighborhoods had some level of access to dockless bicycles but those
with higher incomes and more residents of higher education had more
bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
neighborhoods in New York City with higher affluence had the greatest
proportion of Citi Bike stations.

Chen et al.~(2019) distinguish two types of equity related to bike
share, horizontal and vertical, based on the work of Delbosc and Currie
(Delbosc and Currie, 2011). Horizontal equity leads to balanced or equal
distribution of accessibility and costs for all similar groups, while
vertical equity would involve greater and targeted access for
under-represented or disadvantaged populations (see Chen et al., 2019).
Both are of interest to researchers and transportation planners since
they are often linked in that advantage, or conversely disadvantage, has
spatial patterns. Bike share equity programs, as identified by McNeil et
al.~(2019), can be considered efforts to improve vertical equity since
they favour groups that have benefited less from PBSPs.

@@

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location of stations matters. Kabra et al.~(2020) found that the
majority of bike share usage in Paris comes from areas within 300 m of
stations, which amounts to 2-4 minutes walking by an adult who does not
have a disability. Similar to other public amenities affected by
crowding, the utility of stations is also limited by the \emph{maximum}
number of bicycles that they can hold (e.g., their size). Accessibility
analyses for PBSPs constitute a positive and evaluation-based approach
that also has the potential to inform equity efforts. For instance, Wang
and Lindsey (2019b) investigated whether new or relocated bicycle share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

@@

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bicycle share equity has typically
used a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bicycle share demand (J.
Wang and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation''.

@@

Our analysis builds upon a previous study (Hosford and Winters, 2018),
which found that areas in Hamilton with less advantage are better served
by the city's PBSP compared to other Canadian cities (i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
deprived have greater access. Hosford and Winters (2018, pp.~)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton by default covers more of the
less advantaged areas in the city. However, there is also a great deal
of variation in income in the city center because of the local
university and increasing gentrification. Hosford and Winters (2018)
took a macro-level approach in their analysis by using dissemination
areas \emph{across} the city as the unit of analysis. They did not focus
specifically on the core service area and did not differentiate between
conventional and equity stations.

@@

\begin{verbatim}
## Source : http://tile.stamen.com/toner-lite/8/70/93.png
\end{verbatim}

\begin{verbatim}
## Source : http://tile.stamen.com/toner-lite/8/71/93.png
\end{verbatim}

\begin{verbatim}
## Source : http://tile.stamen.com/toner-lite/8/70/94.png
\end{verbatim}

\begin{verbatim}
## Source : http://tile.stamen.com/toner-lite/8/71/94.png
\end{verbatim}

\begin{verbatim}
## Coordinate system already present. Adding new coordinate system, which will replace the existing one.
\end{verbatim}

@@

The BFCA method was developed to address issues with demand and supply
inflation that result from the overlapping catchment areas produced by
earlier FCA methods (Paez et al., 2019). Paez et al.~(2019) adjusted the
impedance weights so that both supply and demand are proportionally
allocated. The result is a FCA method that balances the population and
level of service by eliminating the over-counting of population and
level of service that leads to distortions in demand and supply.

@@

\begin{verbatim}
## Using plotunit = 'm'
\end{verbatim}

\begin{verbatim}
## Warning in f(...): True north is not meaningful without coord_sf()
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-3}
\end{figure}

\begin{verbatim}
## Using plotunit = 'm'
\end{verbatim}

\begin{verbatim}
## Warning in f(...): True north is not meaningful without coord_sf()
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-5}
\end{figure}

\begin{verbatim}
## Using plotunit = 'm'
\end{verbatim}

\begin{verbatim}
## Warning in f(...): True north is not meaningful without coord_sf()
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-10}
\end{figure}

\begin{verbatim}
## Using plotunit = 'm'
\end{verbatim}

\begin{verbatim}
## Warning in f(...): True north is not meaningful without coord_sf()
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-15}
\end{figure}

@@

Using disaggregated population data, we examined equity in accessibility
to Hamilton Bike Share, with a particular focus on assessing the
contribution of the program's equity stations. Using a BFCA approach,
combined with pycnophylactic interpolation, enabled us to measure
accessibility on a micro scale which better reflects the friction
{[}\emph{is this the right word?}{]} of walking thresholds and avoids
the ``absorption of disparities'', as articulated by Chen et al.~(2019)
and elsewhere in the transportation literature (Rowangould et al.,
2016). The BFCA method also considers demand and supply in the
calculation of accessibility, which is important for services like PBSPs
but not accounted for in other common accessibility models. Unlike the
2SFCA method, there was no inflation or deflation in our accessibility
estimates because population and level of service were preserved (see
Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}). This
differentiates our analysis from similar papers exploring equity in
PBSPs that use larger geographical units of analysis or that focus on
station location instead of level of service. In this way, our paper has
made an important contribution by applying an intuitive and useful
approach to measure accessibility to a PBSP. This open and reproducible
research could be replicated for additional transportation and
statistical analysis using other individual-level data in Hamilton or
revised for PBSPs in other North American cities.

The sensitivity analysis revealed that accessibility to docking stations
is maximized at five minutes and decreases significantly by eight
minutes {[}see Table \ref{tab:accessibility-income}{]}. This reflects
the normative guide advertised on some docking stations in Hamilton
showing other stations within a five minute walk, as well as the
directive of NACTO (City Transportation Officials, 2015). We find that
over 118,000 people can access a bicycle share station within a 15
minute walk, which represents roughly 85\% of the total population in
the core service area {[}see Table \ref{tab:accessibility-income}{]}. At
a minimum threshold of three minutes, too few people can reach stations
which leads to relatively high levels of service since there is little
crowding for the stations. However, accessibility is at its lowest after
eight minutes whereby congestion effects due to increased potential
demand kick in. Chen et al.~(2019) also found that accessibility
decreases as larger geographical units are used in the estimation, since
more people can access the service.

@@

This paper highlights the importance of both the location and size of
docking stations for increasing equity in accessibility to PBSPs.
Several studies have examined who can access PBSPs based on the location
of stations (Babagoli et al., 2019; Qian and Jaller, 2020; Qian et al.,
2020; J. Wang and Lindsey, 2019b), but the potential supply of bicycles
at stations in disadvantaged communities is not considered in the
literature. The size of stations (i.e., potential supply of bicycles)
was also not identified by PBSP owners and operators as a factor that
needed to be considered in equity efforts (Howland et al., 2017). PBSPs
may not necessarily reduce inequities in access if stations are easy to
reach for disadvantaged groups but offer only a small number of
bicycles. Likewise, more people may not opt to use the program if the
supply of bicycles available at nearby stations is insufficient to meet
demand. Babgoli et al.~(2019) found a slight but not statistically
significant increase in the proportion of neighborhoods with the highest
levels of poverty with stations after the Citi Bike expansion in 2015.
Although the Citi Bike expansion was not specifically driven by a desire
to reduce inequity in access, 16\% of neighborhoods with the highest
levels of poverty had stations compared to 12\% before. This leaves open
the question as to whether the stations held enough bicycles to meet
demand, and whether congestion effects in neighbourhoods with more
poverty are reducing the potential of bike share trips.

The following policy and planning implications can inform efforts for
increasing equity in Hamilton Bike Share, and potentially other PBSPs in
North America. Our analysis has identified specific areas that have both
low accessibility and low median total household income. Figures
\ref{fig:figure-bi-map-threshold-3},
\ref{fig:figure-bi-map-threshold-5},
\ref{fig:figure-bi-map-threshold-10}, and
\ref{fig:figure-bi-map-threshold-15} highlight potential locations for
new equity stations to better accommodate groups with lower
socioeconomic status. Additional stations are needed to further reduce
inequities in accessibility. This information will be useful to
transportation planners in Hamilton and HBS operators. Our findings also
suggest that PBSPs should consider increasing the size of stations in
disadvantaged communities. Docking stations added to target vertical
equity should have an adequate supply of public bicycles, particularly
more in areas that have a strong interest in adopting bike share.
Finally, distance to stations is associated with bike share use (Fuller
et al., 2011; J. Wang and Lindsey, 2019b), especially when the trip
starts at the residence (Fishman et al., 2013), and can be a barrier to
using PBSPs (Fishman et al., 2014). To address this issue,
transportation planners and PBSP operators should consult with groups
and communities who are under-represented among bike share users to
determine how far they are willing to walk to reach a docking station
and to get their input on the location of new docking stations.

@@

This paper did not examine or compare ridership data between
conventional and equity stations. Therefore, further research is needed
to determine whether the addition of equity stations encouraged more
bicycling for disadvantaged groups living near them. Other studies have
specifically looked at differences in trip type, frequency, or length
among users from disadvantaged neighborhoods (Caspi and Noland, 2019;
Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), but our analysis is
limited by the lack of publicly available route and individual user data
to conduct similar analyses for HBS.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining equity in accessibility to bike share - a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

@@

The potential of public bicycle share programs (PBSPs) to increase
bicycling levels is but one of many reasons for implementing such
programs in urban areas (Hosford et al., 2019, see 2018). As a healthy,
inexpensive, and convenient form of public transportation, shared
bicycles can encourage individuals to take up bicycling for short local
trips or first and last mile trips to other public transportation
instead of using personal vehicles. These programs can also play a role
in advancing transportation equity if they make bicycling more
accessible to disadvantaged populations. Although PBSPs are available to
the general public in over 800 cities worldwide (Fishman, 2016), and
ought to be accessible to any individual who wishes to use them,
research on PBSPs indicates that inequities persist with respect to who
can use and access them.

@@

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users
(Fishman, 2016). In many cities, males use bike share more than females
(Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
Reilly et al., 2020b; Winters et al., 2019) as do younger age cohorts
(Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011). However,
one study found that bike share users in Washington, DC were more likely
to be female (Buck et al., 2013), which suggests that the gender gap
among bicyclists who use PBSPs is less disparate than the gap for
personal bicycle use (Fishman, 2016). There is some evidence that bike
share users are less likely to own a car (Buck et al., 2013; Reilly et
al., 2020a). However, the relationship between income or education and
bike share use is less clear-cut. Stations in disadvantaged communities
in Chicago have been found to generate most of the average annual trips
(Qian and Jaller, 2020) and individuals from minority or lower
socioeconomic status neighborhoods in Minneapolis-St.~Paul also used the
city's PBSP more (Wang and Lindsey, 2019a). Similar findings were
reported in London (Ogilvie and Goodman, 2012). Being university
educated was a significant correlate of bike share use in Montreal,
Canada (Fuller et al., 2011). Not coincidentally, financial savings have
been found to motivate those on a low income to use bike share (Fishman,
2016).

@@

The introduction of PBSPs has been accompanied by a flurry of research
focusing on benefits from them. Many studies have examined differences
in demographics or socioeconomic status between those who use or have
access to PBSPs and those who don't, while other research has explored
spatial inequities in where stations are located (e.g., Hosford and
Winters, 2018; Hull Grasso et al., 2020; Mooney et al., 2019; Qian and
Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian and Jaller
(2020) estimated ridership in the city's PBSP and found that a minority
of bike share stations were located in disadvantaged communities, while
annual members from such areas had a lower share of trips compared to
other areas in the city. Similar results were found in Philadelphia.
Despite efforts to increase equity within the city's PBSP, census block
groups with lower median income generated fewer trips (Caspi and Noland,
2019). Trips from stations in such areas were utilitarian (e.g.,
commuting to work), which points to the importance of ensuring equitable
access (Caspi and Noland, 2019). In the case of Seattle, all
neighborhoods had some level of access to dockless bicycles but those
with higher incomes and more residents of higher education had more
bikes (Mooney et al., 2019). Babagoli et al. (2019) also found that
neighborhoods in New York City with higher affluence had the greatest
proportion of Citi Bike stations.

Chen et al. (2019) distinguish two types of equity related to bike
share, horizontal and vertical, based on the work of Delbosc and Currie
(Delbosc and Currie, 2011). Horizontal equity leads to balanced or equal
distribution of accessibility and costs for all similar groups, while
vertical equity would involve greater and targeted access for
under-represented or disadvantaged populations (see Chen et al., 2019).
Both are of interest to researchers and transportation planners since
they are often linked in that advantage, or conversely disadvantage, has
spatial patterns. Bike share equity programs, as identified by McNeil et
al. (2019), can be considered efforts to improve vertical equity since
they favour groups that have benefited less from PBSPs.

@@

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location of stations matters. Kabra et al. (2020) found that the
majority of bike share usage in Paris comes from areas within 300 m of
stations, which amounts to 2-4 minutes walking by an adult who does not
have a disability. Similar to other public amenities affected by
crowding, the utility of stations is also limited by the \emph{maximum}
number of bicycles that they can hold (e.g., their size). Accessibility
analyses for PBSPs constitute a positive and evaluation-based approach
that also has the potential to inform equity efforts. For instance, Wang
and Lindsey (2019b) investigated whether new or relocated bicycle share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

@@

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bicycle share equity has typically
used a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bicycle share demand (Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al. (2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation.''

@@

Our analysis builds upon a previous study (Hosford and Winters, 2018),
which found that areas in Hamilton with less advantage are better served
by the city's PBSP compared to other Canadian cities (i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
deprived have greater access. Hosford and Winters (2018, pp.~)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas.'' Therefore, the
core service area for the PBSP in Hamilton by default covers more of the
less advantaged areas in the city. However, there is also a great deal
of variation in income in the city center because of the local
university and increasing gentrification. Hosford and Winters (2018)
took a macro-level approach in their analysis by using dissemination
areas \emph{across} the city as the unit of analysis. They did not focus
specifically on the core service area and did not differentiate between
conventional and equity stations.

@@

@@

The BFCA method was developed to address issues with demand and supply
inflation that result from the overlapping catchment areas produced by
earlier FCA methods (Paez et al., 2019). Paez et al. (2019) adjusted the
impedance weights so that both supply and demand are proportionally
allocated. The result is a FCA method that balances the population and
level of service by eliminating the over-counting of population and
level of service that leads to distortions in demand and supply.

@@

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-3}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-5}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-10}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-15}
\end{figure}

@@

Using disaggregated population data, we examined equity in accessibility
to Hamilton Bike Share, with a particular focus on assessing the
contribution of the program's equity stations. Using a BFCA approach,
combined with pycnophylactic interpolation, enabled us to measure
accessibility on a micro scale which better scale at which walking takes
place and avoids the ``absorption of disparities,'' as articulated by
Chen et al. (2019) and elsewhere in the transportation literature
(Rowangould et al., 2016). The BFCA method also considers demand and
supply in the calculation of accessibility, which is important for
services like PBSPs but not accounted for in other common accessibility
models. Unlike the 2SFCA method, there was no inflation or deflation in
our accessibility estimates because population and level of service were
preserved (see Figure \ref{fig:figure-10} and Figure
\ref{fig:figure-11}). This differentiates our analysis from similar
papers exploring equity in PBSPs that use larger geographical units of
analysis or that focus on station location instead of level of service.
In this way, our paper has made an important contribution by applying an
intuitive and useful approach to measure accessibility to a PBSP. This
open and reproducible research could be replicated for additional
transportation and statistical analysis using other individual-level
data in Hamilton or revised for PBSPs in other North American cities.

The sensitivity analysis revealed that accessibility to docking stations
is maximized at five minutes and decreases significantly by eight
minutes {[}see Table \ref{tab:accessibility-income}{]}. This reflects
the normative guide advertised on some docking stations in Hamilton
showing other stations within a five minute walk, as well as the
directive of NACTO (City Transportation Officials, 2015). We find that
over 118,000 people can access a bicycle share station within a 15
minute walk, which represents roughly 85\% of the total population in
the core service area {[}see Table \ref{tab:accessibility-income}{]}. At
a minimum threshold of three minutes, too few people can reach stations
which leads to relatively high levels of service since there is little
crowding for the stations. However, accessibility is at its lowest after
eight minutes whereby congestion effects due to increased potential
demand kick in. Chen et al. (2019) also found that accessibility
decreases as larger geographical units are used in the estimation, since
more people can access the service.

@@

This paper highlights the importance of both the location and size of
docking stations for increasing equity in accessibility to PBSPs.
Several studies have examined who can access PBSPs based on the location
of stations (Babagoli et al., 2019; Qian et al., 2020; Qian and Jaller,
2020; Wang and Lindsey, 2019b), but the potential supply of bicycles at
stations in disadvantaged communities is not considered in the
literature. The size of stations (i.e., potential supply of bicycles)
was also not identified by PBSP owners and operators as a factor that
needed to be considered in equity efforts (Howland et al., 2017). PBSPs
may not necessarily reduce inequities in access if stations are easy to
reach for disadvantaged groups but offer only a small number of
bicycles. Likewise, more people may not opt to use the program if the
supply of bicycles available at nearby stations is insufficient to meet
demand. Babgoli et al. (2019) found a slight but not statistically
significant increase in the proportion of neighborhoods with the highest
levels of poverty with stations after the Citi Bike expansion in 2015.
Although the Citi Bike expansion was not specifically driven by a desire
to reduce inequity in access, 16\% of neighborhoods with the highest
levels of poverty had stations compared to 12\% before. This leaves open
the question as to whether the stations held enough bicycles to meet
demand, and whether congestion effects in neighbourhoods with more
poverty are reducing the potential of bike share trips.

The following policy and planning implications can inform efforts for
increasing equity in Hamilton Bike Share, and potentially other PBSPs in
North America. Our analysis has identified specific areas that have both
low accessibility and low median total household income. Figures
\ref{fig:figure-bi-map-threshold-3},
\ref{fig:figure-bi-map-threshold-5},
\ref{fig:figure-bi-map-threshold-10}, and
\ref{fig:figure-bi-map-threshold-15} highlight potential locations for
new equity stations to better accommodate groups with lower
socioeconomic status. Additional stations are needed to further reduce
inequities in accessibility. This information will be useful to
transportation planners in Hamilton and HBS operators. Our findings also
suggest that PBSPs should consider increasing the size of stations in
disadvantaged communities. Docking stations added to target vertical
equity should have an adequate supply of public bicycles, particularly
more in areas that have a strong interest in adopting bike share.
Finally, distance to stations is associated with bike share use (Fuller
et al., 2011; Wang and Lindsey, 2019b), especially when the trip starts
at the residence (Fishman et al., 2013), and can be a barrier to using
PBSPs (Fishman et al., 2014). To address this issue, transportation
planners and PBSP operators should consult with groups and communities
who are under-represented among bike share users to determine how far
they are willing to walk to reach a docking station and to get their
input on the location of new docking stations.

@@

This paper did not examine or compare ridership data between
conventional and equity stations. Therefore, further research is needed
to determine whether the addition of equity stations encouraged more
bicycling for disadvantaged groups living near them. Other studies have
specifically looked at differences in trip type, frequency, or length
among users from disadvantaged neighborhoods (Caspi and Noland, 2019;
Qian and Jaller, 2020; Wang and Lindsey, 2019a), but our analysis is
limited by the lack of publicly available route and individual user data
to conduct similar analyses for HBS.
